The Heavy Photon Search Test Detector

### 2013-11-08 hps-testrun-nim.tex
Adding new kinematic plots from TestRun-v6 geometry. Removing a few of the MCS distributions. Removed plot of G4 comparison, just use table to make the point.

By selecting \ee{} pairs from the triggered events we are able to study basic distributions of pair production 
kinematics and in particular those related to our vertex performance. Pairs of opposite charge tracks, one in the 
top and one in the bottom half of the SVT, with momentum larger than 400~MeV were selected. 
The pair production 
kinematics are relatively well reproduced given the alignment of the tracker; Fig.~\ref{fig:pair_kin} shows the 
invariant mass and ratio of electron momentum over the sum of electron and positron. 
 \begin{center}
{\small
\begin{figure*}[ht]
   \includegraphics[width=5cm]{figures/h_trk_top_px_h_trk_top_px_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
   \includegraphics[width=5cm]{figures/h_invM_h_invM_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
   \includegraphics[width=5cm]{figures/h_sumE_h_sumE_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
%   \includegraphics[scale=0.25]{test2012/vertexing/figures/h_ratioEsum_h_ratioEsum_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
\caption{Kinematic distributions for e$^+$e$^-$ pairs selected by opposite charged tracks in the top and bottom half of the tracker: track momentum in the top half of the SVT (left), invariant mass (middle) and the sum of the track momentum for the pair.}
\label{fig:pair_kin}
\end{figure*}
}
\end{center}

For the vertexing 
performance the foremost difference compared to electron beam running is that the target was 
located approximately 67~cm from our nominal target position; giving almost collinear tracks in the detector. 
This degrades the vertex resolution along the 
beam line compared to that expected in an electron beam with tracks from the nominal target position. 
Furthermore, tails of the vertex distributions are impossible to study with the finite data sample obtained 
in the test run. Nevertheless, useful information can still be 
obtained by studying the vertex distributions. Figure~\ref{fig:vtx_pos} shows the distance of closest 
approach of the momentum vectors extrapolated in the 
upstream direction from our analyzing magnet, taking into account the measured fringe field of the PS magnet. 
 \begin{center}
{\small
	 \begin{figure*}[ht]
	\includegraphics[width=5cm]{figures/h_vtx_fr_x_h_vtx_x_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\includegraphics[ width=5cm]{figures/h_vtx_fr_y_h_vtx_y_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\includegraphics[width=5cm]{figures/h_vtx_fr_z_h_vtx_z_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
	\caption{{\bf \textcolor{red}{NEED TO BE UPDATED}}Vertex position represented by the distance of 
	closest approach of the extrapolated momentum vectors upstream of the PS magnet taking into account 
	the measured fringe field. The overall shift from zero in the x-direction is 
	due to a 30~mrad rotation of the SVT with respect to the beam line.}
	\label{fig:vtx_pos}
\end{figure*}
}
\end{center}
While the tails of the vertex distribution expected in electron beam running is not accessible here the 
fact that the core is relatively well described provides confidence of the description of the amount of 
material and the model used for simulating multiple scattering in the target.  
These are crucial for benchmarking the HPS \Aprime{} physics reach since both the mass and vertex resolution 
that determine the sensitivity are limited by multiple scattering.

@@

One of the main goals of the HPS Test was to evaluate the description of the tails of the MCS
to gain further confidence in the expected detector occupancy in the full HPS experiment. 
Previous studies of multiple scattering angles show good agreement with the \moliere{} 
theory~\cite{Moliere:1948zz} for a wide range 
of materials and projectiles~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467}. We have verified that 
the angular distribution $F(\theta)$ in the differential cross section $d\sigma = F( \theta) d(\cos\theta) d\phi$ 
for the \egs{}~\cite{egs5} simulation program show good agreement with \moliere{}'s analytical formula as 
formulated by Bethe~\cite{Bethe:1953va}. 
The small angle approximation was also shown to be in agreement with the theory formulated by Gaudsmit and 
Saunderson~\cite{Goudsmit:1940zza,Goudsmit:1940zz} that is valid for any angle. 
While \egs{} uses the more complex and time consuming \moliere{} 
formula, \geant{} uses an approximation with two, continuously joined, functions to take into account 
small and large angle scattering. Due to the explicit function in large angle approximation used by \geant{} 
we expect that \geant{} will overestimate the angular distribution at angles larger than a few mrad.

@@

We measure the vertical angular distributions of electrons and positrons using the ECal. 
Cluster reconstruction and calibration was performed as described in Sec.~\ref{sec:ecal}. The measured 
angular distribution in the ECal for the three converter thicknesses are shown in 
Fig.~\ref{fig:ang_distr_data} (left).  
\begin{figure}[]
\begin{center}
{\small
	\includegraphics[ width=7cm]{figures/rate_ecalrow_raw.png}
	\includegraphics[ width=7cm]{figures/rate_ecalrow_norm_subtr.png}
	\caption{Measured raw vertical angular distributions before (top) and after (bottom) 
	normalization and background subtraction.}
	\label{fig:ang_distr_data}
}
\end{center}
\end{figure}
The background fraction for the three converter thicknesses was 16\%, 52\% and 71\% 
for converter thicknesses of 1.6\%, 0.45\% and 0.18\%, respectively. The background fraction was also 
cross-checked by pointing back tracks reconstructed in the SVT to identify the fraction of tracks not originating 
from the converter. We also checked that the contribution from photons to our triggered 
sample was much less than 2\%. These measured angular distributions are compared to simulation to validate 
the modeling of the MCS. \egs{}~\cite{egs5} is used to generate 
the electromagnetic interactions in the converter while \geant{} is used to simulate the particles after the 
converter.  Figure~\ref{fig:ang_distr_dataMC} shows the angular distribution comparing data and 
\egs{} normalized to 1~s of beam at 90~nA beam current. Taking into account systematics uncertainties 
described below reasonable agreement is obtained;   total rates for each converter thickness is shown in 
Fig.~\ref{fig:rate_vs_thickness} and summarized in Tab.~\ref{tab:ang_distr_dataMC}.
\begin{center}
{\small
	\begin{figure*}[t]
	\includegraphics[ scale=0.25]{figures/dataMC_1351_Hit_Y_top_norm_bkgsub.png}
	\includegraphics[ scale=0.25]{figures/dataMC_1354_Hit_Y_top_norm_bkgsub.png}
	\includegraphics[ scale=0.25]{figures/dataMC_1353_Hit_Y_top_norm_bkgsub.png}
	\caption{Comparison between the observed and predicted angular 
distribution using \egs{} for a converter thickness of 1.6\% (left), 0.45\% (middle) and 0.18\% 
(right).  Only statistical uncertainties are shown. }
	\label{fig:ang_distr_dataMC}
	\end{figure*}
}
\end{center}
\begin{center}
{\small
	\begin{figure*}[t]
	\includegraphics[ scale=0.3]{figures/dataMC_geant4.png}
	\includegraphics[ scale=0.3]{figures/dataMC_egs.png}
	\caption{The measured rate as a function of converter thickness comparing \geant{} (left) 
	and \egs{} (right).} 
	\label{fig:rate_vs_thickness}
	\end{figure*}
}
\end{center}
The total systematic uncertainty was estimated to be between 10-18\% depending on the run, they include:  
uncertainty on the integrated current normalization (5\%), alignment of the ECal, uncertainty from the 
background normalization and limited Monte Carlo statistics.  
\begin{table}
\begin{center}
{\small
\begin{tabular}{|l|c|c|c|}
\hline
Converter (\% $X_0$) & 1.60 & 0.45 &	0.18 \\
\hline
\egs{} &	1162 $\pm$ 112 &	255 $\pm$ 28 &	94 $\pm$ 17	\\
\hline
\geant{} & 2633 $\pm$ 250 & 	371 $\pm$ 38 &	114 $\pm$ 18 \\
\hline
Observed 	& 1064 $\pm$ 2 & 196 $\pm$ 1 &	92 $\pm$ 1 \\						
\hline
\end{tabular}
\caption{ Observed and predicted number of events for 1~s of beam at 90~nA for three different converter 
thicknesses. The uncertainty on the prediction includes systematic uncertainties. }
\label{tab:ang_distr_dataMC}
}
\end{center}
\end{table}

>>>>

By selecting \ee{} pairs from the triggered events we are able to study basic distributions of pair production 
kinematics and in particular those related to our vertex performance. Pairs of opposite charge tracks, one in the 
top and one in the bottom half of the SVT, with momentum larger than 400~MeV were selected. 
The pair production 
kinematics are relatively well reproduced given the alignment of the tracker; Fig.~\ref{fig:pair_kin} shows the 
invariant mass and ratio of electron momentum over the sum of electron and positron. 
 \begin{center}
{\small
\begin{figure*}[ht]
   \includegraphics[width=5cm]{figures/h_trk_top_px_h_trk_top_px_trigsel4hit_pair1351_twotrkfilt-v6-paper}
   \includegraphics[width=5cm]{figures/h_invM_h_invM_trigsel4hit_pair1351_twotrkfilt-v6-paper}
   \includegraphics[width=5cm]{figures/h_sumE_h_sumE_trigsel4hit_pair1351_twotrkfilt-v6-paper}
%   \includegraphics[scale=0.25]{test2012/vertexing/figures/h_ratioEsum_h_ratioEsum_dataMC_trigseltwotrksel4hit_recoilmc_twotrkfilt.png}
\caption{Kinematic distributions for e$^+$e$^-$ pairs selected by opposite charged tracks in the top and bottom half of the tracker: track momentum in the top half of the SVT (left), invariant mass (middle) and the sum of the track momentum for the pair.}
\label{fig:pair_kin}
\end{figure*}
}
\end{center}

For the vertexing 
performance the foremost difference compared to electron beam running is that the target was 
located approximately 67~cm from our nominal target position; giving almost collinear tracks in the detector. 
This degrades the vertex resolution along the 
beam line compared to that expected in an electron beam with tracks from the nominal target position. 
Furthermore, tails of the vertex distributions are impossible to study with the finite data sample obtained 
in the test run. Nevertheless, useful information can still be 
obtained by studying the vertex distributions. Figure~\ref{fig:vtx_pos} shows the distance of closest 
approach of the momentum vectors extrapolated in the 
upstream direction from our analyzing magnet, taking into account the measured fringe field of the PS magnet. 
 \begin{center}
{\small
	 \begin{figure*}[ht]
	% Plot info:
	% TestRun-v6 geometry, twotrkfilter, trigsel, hps-java-1.8-SNAPSHOT, 4hits, chi2<10
	\includegraphics[width=5cm]{figures/h_vtx_fr_x_h_vtx_x_trigsel4hit_pair1351_twotrkfilt-v6-paper}
	\includegraphics[width=5cm]{figures/h_vtx_fr_y_h_vtx_y_trigsel4hit_pair1351_twotrkfilt-v6-paper}
	\includegraphics[width=5cm]{figures/h_vtx_fr_z_h_vtx_z_trigsel4hit_pair1351_twotrkfilt-v6-paper}
	\caption{ Vertex position represented by the distance of 
	closest approach of the extrapolated momentum vectors upstream of the PS magnet taking into account 
	the measured fringe field. The overall shift from zero in the x-direction is 
	due to a 30~mrad rotation of the SVT with respect to the beam line.}
	\label{fig:vtx_pos}
\end{figure*}
}
\end{center}
While the tails of the vertex distribution expected in electron beam running is not accessible here the 
fact that the core is relatively well described provides confidence of the description of the amount of 
material and the model used for simulating multiple scattering in the target.  
These are crucial for benchmarking the HPS \Aprime{} physics reach since both the mass and vertex resolution 
that determine the sensitivity are limited by multiple scattering.

@@

One of the main goals of the HPS Test was to evaluate the description of the tails of the MCS
to gain further confidence in the expected detector occupancy in the full HPS experiment. 
Previous studies of multiple scattering angles show good agreement with the \moliere{} 
theory~\cite{Moliere:1948zz} for a wide range 
of materials and projectiles~\cite{Attwood:2005zz,Shen:1978ha,Gottschalk1993467}. We have verified that 
the angular distribution $F(\theta)$ in the differential cross section $d\sigma = F( \theta) d(\cos\theta) d\phi$ 
for the \egs{}~\cite{egs5} simulation program show good agreement with \moliere{}'s analytical formula as 
formulated by Bethe~\cite{Bethe:1953va}. 
The small angle approximation was also shown to be in agreement with the theory formulated by Gaudsmit and 
Saunderson~\cite{Goudsmit:1940zza,Goudsmit:1940zz} that is valid for any angle. 
While \egs{} uses the more complex and time consuming \moliere{} 
formula, the default physics list of \geant{} uses the so-called Urban model, an approximation with two, 
continuously joined, functions to take into account 
small and large angle scattering. Due to the explicit function in large angle approximation used by \geant{} 
we expect that \geant{} will overestimate the angular distribution at angles larger than a few mrad.

@@

We measure the vertical angular distributions of electrons and positrons using the ECal. 
Cluster reconstruction and calibration was performed as described in Sec.~\ref{sec:ecal}. The measured 
angular distribution in the ECal for the three converter thicknesses are shown in 
Fig.~\ref{fig:ang_distr_data} (left). The data has been normalized to the integrated beam current and 
the upstream background has been subtracted. 
\begin{figure}[]
\begin{center}
{\small
%	\includegraphics[ width=7cm]{figures/rate_ecalrow_raw.png}
	\includegraphics[ width=7cm]{figures/rate_ecalrow_norm_subtr.png}
	\caption{Measured vertical angular distributions before after integrated beam current 
	normalization and upstream background subtraction.}
	\label{fig:ang_distr_data}
}
\end{center}
\end{figure}
The background fraction for the three converter thicknesses was 16\%, 52\% and 71\% 
for converter thicknesses of 1.6\%, 0.45\% and 0.18\%, respectively. The background fraction was also 
cross-checked by pointing back tracks reconstructed in the SVT to identify the fraction of tracks not originating 
from the converter. We also checked that the contribution from photons to our triggered 
sample was much less than 2\%. These measured angular distributions are compared to simulation to validate 
the modeling of the MCS. \egs{}~\cite{egs5} is used to generate 
the electromagnetic interactions in the converter while \geant{} is used to simulate the particles after the 
converter.  Figure~\ref{fig:ang_distr_dataMC} shows the angular distribution comparing data and 
\egs{} normalized to 1~s of beam at 90~nA beam current with a converter thickness of 1.6\%. 
\begin{center}
{\small
	\begin{figure}[]
%	\begin{figure*}[t]
	\includegraphics[ scale=0.25]{figures/dataMC_1351_Hit_Y_top_norm_bkgsub.png}
%	\includegraphics[ scale=0.25]{figures/dataMC_1354_Hit_Y_top_norm_bkgsub.png}
%	\includegraphics[ scale=0.25]{figures/dataMC_1353_Hit_Y_top_norm_bkgsub.png}
	\caption{Comparison between the observed and predicted angular distribution using \egs{} for a 
	converter thickness of 1.6\%. Similar agreement is found for 0.45\% and 0.18\% converter thicknesses. 
	Only statistical uncertainties are shown. }
	\label{fig:ang_distr_dataMC}
%	\end{figure*}
	\end{figure}
}
\end{center}
Reasonable agreement is obtained in the most important region close to the beam. There is a hint of a 
slightly different slope of the data at larger angles; this difference is covered by systematic uncertainties.  The 
total rates for each converter thickness is shown in Fig.~\ref{fig:rate_vs_thickness} and summarized in 
Tab.~\ref{tab:ang_distr_dataMC}.
\begin{center}
{\small
	\begin{figure}[]
%	\begin{figure*}[]
	%\includegraphics[ scale=0.3]{figures/dataMC_geant4.png}
	\includegraphics[ scale=0.3]{figures/dataMC_egs.png}
	\caption{The measured rate as a function of converter thickness for data and \egs{} simulation of MCS in 	the converter.} 
	\label{fig:rate_vs_thickness}
%	\end{figure*}
	\end{figure}
}
\end{center}
The total systematic uncertainty was estimated to be between 10-18\% depending on the run, they include:  
uncertainty on the integrated current normalization (5\%), alignment of the ECal, uncertainty from the 
background normalization and limited Monte Carlo statistics.  
\begin{table}
\begin{center}
{\small
\begin{tabular}{|l|c|c|c|}
\hline
Converter (\% $X_0$) & 1.60 & 0.45 &	0.18 \\
\hline
\egs{} &	1162 $\pm$ 112 &	255 $\pm$ 28 &	94 $\pm$ 17	\\
\hline
\geant{} & 2633 $\pm$ 250 & 	371 $\pm$ 38 &	114 $\pm$ 18 \\
\hline
Observed 	& 1064 $\pm$ 2 & 196 $\pm$ 1 &	92 $\pm$ 1 \\						
\hline
\end{tabular}
\caption{ Observed and predicted number of events for 1~s of beam at 90~nA for three different converter 
thicknesses. The uncertainty on the prediction includes systematic uncertainties. }
\label{tab:ang_distr_dataMC}
}
\end{center}
\end{table}
As expected \egs{} show good agreement with our data while \geant{} using the default physics list 
overestimates the MCS at these large angles.
